A Monte Carlo Synthetic Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-07-16 doc/jcomp2012/jcomp2012.tex
updated latex and makefile for figures

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{tmadd,tmath}
\usepackage[mathcal]{euscript}

@@

  %% Title, authors and addresses

@@

  %%---------------------------------------------------------------------------%%
  \begin{abstract}

@@

  %%---------------------------------------------------------------------------%%
  \begin{keyword}
    %% keywords here, in the form: keyword \sep keyword
    radiation diffusion \sep Monte Carlo \sep sparse matrix systems
    %% MSC codes here, in the form: \MSC code \sep code
    %% or \MSC[2008] code \sep code (2000 is the default)

  \end{keyword}

@@

%%
%% Start line numbering here if you want
%%
% \linenumbers

@@

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for 3D, time-dependent discrete systems.  In this
note, we extend our previous work in the following manner:
\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this solver to the 1T thermal radiation diffusion
  equation in 3D.
\end{itemize}
In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
our new synthetic-acceleration scheme and connect with previous work
by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The 1T
radiation diffusion equation and its discrete form will be derived in
\S~\ref{sec:discr-form-radi}.  We discuss using Synthetic-Accleration
Monte Carlo to solve the discrete radiation diffusion equation in
\S~\ref{sec:solut-radi-diff}.  Results from two 3D test problems are
given in \S~\ref{sec:results}.  We finish with conclusions and ideas
for future work in \S~\ref{sec:conclusions}.

@@

Concisely, the Sequential Monte Carlo method solves
Eq.~(\ref{eq:Ax=b}) using the adjoint solution technique described in
\S~\ref{sec:adjoint-method}.  The following iteration scheme is
applied:
\begin{subequations}
  \begin{gather}
    \vr^{l+1} = \vb - \vA\vx^{l}\:, \\
    \vA\delta\vx^{l+1} = \vr^{l+1}\:, \label{eq:residual_solve}\\
    \vx^{l+1} = \vx^{l} + \delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
In this scheme, the adjoint method is used to estimate the solution to
Eq.~(\ref{eq:residual_solve}), and the residual is iterated to
convergence.  This iteration sequence is closely related to iterative
refinement; the exception being that there is no update to to
$x^{l+1}$ between residual iterations.

@@

\subsection{Justification of Adjoint Error Estimation}

@@

Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each case
to converge in seconds and Fig.~\ref{fig:poisson_iterations} gives the
number of iterations needed for each case to converge. We see clearly
in Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint solver
with MCSA results in several orders of magnitude speedup over the
direct solver while the number of iterations required to converge is
of a similar scale. We expect this for several reasons. First, with an
equivalent number of histories specified for both solvers and a system
of size $N$, the direct solver will compute $N$ times the number of
histories random walks while the adjoint solver will only compute the
number of histories specified. Initially, this may not seem like a
fair comparison, however, we see from Fig.~\ref{fig:poisson_cpu_time}
that the number of iterations requried to converge is approximately
the same. This is necessary in the direct method to ensure a
contribution from each state as the random walk sequence will only
contribute to the starting state. For the adjoint method, because the
random sequence contributes to the state in which it currently
resides, fewer histories are neccessary to compute contributions from
all the nonzero parts of the source. Therefore, although the total
number of histories per cycle is in fact larger for the direct method,
the solution computed at each iteration is no better than that
computed by the adjoint method.

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.eps}
  \caption{\sl CPU Time (s) to converge vs. Problem Size (N for an NxN
    problem). Both the adjoint and direct solvers are used with the
    five point and nine point stencils. A significant CPU time speedup
    is noted with the adjoint method due to the reduced number of
    histories.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.eps}
  \caption{\sl Iterations to converge vs. Problem Size (N for an NxN
    problem). Both the adjoint and direct solvers are used with the
    five point and nine point stencils. Both methods give similar
    iteration behavior.}
  \label{fig:poisson_iterations}
\end{figure}

@@

The discrete diffusion equation in (\ref{eq:discrete-diffusion}) can
be written in matrix-operator form as follows,
\begin{equation}
  \ve{D}\bphi = \ve{q}\:,
  \label{eq:diffusion-matrix}
\end{equation}
where $\ve{D}$ is a symmetric, positive-definite (SPD) operator.
Unfortunately, for many problems $\rho(\ve{D}) > 1$.  Therefore, we
must precondition Eq.~(\ref{eq:diffusion-matrix}) in order to use the
MCSA method.  Preconditioning accomplishes two purposes: it reduces
the spectral radius of $\ve{D}$ so that Monte Carlo solution methods
are applicable, and it reduces the number of iterations required for
convergence.  We apply left-preconditioning and write
\begin{equation}
  \ve{M}^{-1}\ve{D}\bphi = \ve{M}^{-1}\ve{q}\:.
  \label{eq:preconditioned-diffusion}
\end{equation}
A good choice for the preconditioner, $\ve{M}$, is
\begin{equation}
  \ve{M}=\diag(\ve{D})\:.
  \label{eq:preconditioner}
\end{equation}
Setting $\ve{M}$ equal to the diagonal elements of $\ve{D}$
accomplishes several objectives:
\begin{itemize}
\item It reduces the spectral radius such that
  $\rho(\ve{M}^{-1}\ve{D}) < \rho(\ve{D})$ and
  $\rho(\ve{M}^{-1}\ve{D}) < 1$.  Thus, convergence is guaranteed and
  fewer iterations are required resulting in accelerated convergence.
\item Makes the diagonal elements of the iteration matrix zero,
  $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no time
  wasted in transitions $i\rightarrow i$ during the random walk
  process.
\end{itemize}
The second item is quite important for Monte Carlo solutions of linear
systems, and this type of scaling should be performed regardless of
any additional preconditioner choices.

@@

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA
method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
method requires two solution steps:
\begin{enumerate}
\item A fixed-point iteration to estimate $\bphi^{l+1/2}$.  This is a
  matrix-vector multiply operation.
\item An adjoint Monte Carlo solve to estimate $\delta\bphi^{l+1/2}$.
\end{enumerate}
For clarity, we rewrite the iteration scheme in
Eqs.~(\ref{eq:general-MCSA}) using the diffusion operator notation
from Eq.~(\ref{eq:preconditioned-diffusion}),
\begin{align*}
  %%
  \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
  \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\
  %%
  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
  \quad & (\text{compute residual})\\
  %%
  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
  (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\
  %%
  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
  (\text{update}\ \bphi)
\end{align*}
As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo
method only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now
describe each of these steps in turn.

First, we use Eq.~(\ref{eq:preconditioned-discrete-diffusion}) to
evaluate $\bphi^{l+1/2}$ via a single fixed-point iteration as
follows:
\begin{equation}
  \begin{aligned}
    \phi_{ijk}^{l+1/2} =  
    \frac{\sigm_{i+1}}{\sigT}\frac{\Delta_{i+1}}{\Di}&\phi_{i+1\,jk}^l
    + \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}^l
    +
    \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}^l \\ 
    &+ \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}^l 
    + \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}^l
    + \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}^l
    + \frac{\qn_{ijk}}{\sigT}\:.
  \end{aligned}
\end{equation}
This result is used to compute the residual of the system.  The
residual then becomes the source for the Monte Carlo simulation.

@@

\begin{figure}[htpb!]
  \begin{center}
    \begin{tabular}{m{2in}m{4in}}
      \subfigure[Edit points]{\includegraphics[width=2in,clip]{points.eps}} &
      \subfigure[Temperature]{\includegraphics[width=4in,clip]{multi_mat.eps}}
    \end{tabular}
  \end{center}
  \caption{(a) Edit points and (b) time-evolution of the temperature
    at each point.  All points are centered in the $z$-axis.}
  \label{fig:edit-points}
\end{figure}

>>>>

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
\usepackage{graphicx}
\usepackage{subfigure}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{tmadd,tmath}
\usepackage[mathcal]{euscript}
\usepackage{color}

@@

@@

  \begin{abstract}

@@

  \begin{keyword}
    radiation diffusion \sep Monte Carlo \sep sparse matrix systems
  \end{keyword}

@@

@@

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for 3D, time-dependent discrete systems.  In this
paper, we extend our previous work in the following manner:
\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this solver to the 1T thermal radiation diffusion
  equation in 3D.
\end{itemize}
In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
our new synthetic-acceleration scheme and connect with previous work
by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The 1T
radiation diffusion equation and its discrete form will be derived in
\S~\ref{sec:discr-form-radi}.  We discuss using Synthetic-Accleration
Monte Carlo to solve the discrete radiation diffusion equation in
\S~\ref{sec:solut-radi-diff}.  Results from two 3D test problems are
given in \S~\ref{sec:results}.  We finish with conclusions and ideas
for future work in \S~\ref{sec:conclusions}.

@@

Concisely, the Sequential Monte Carlo method solves
Eq.~(\ref{eq:Ax=b}) using the adjoint solution technique described in
\S~\ref{sec:adjoint-method}.  The following iteration scheme is
applied:
\begin{subequations}
  \begin{gather}
    \vr^{l+1} = \vb - \vA\vx^{l}\:, \\
    \vA\delta\vx^{l+1} = \vr^{l+1}\:, \label{eq:residual_solve}\\
    \vx^{l+1} = \vx^{l} + \delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
In this scheme, the Monte Carlo adjoint method is used to estimate the
solution to Eq.~(\ref{eq:residual_solve}), and the residual is
iterated to convergence.  This iteration sequence is closely related
to iterative refinement; the exception being that there is no update
to to $x^{l+1}$ between residual iterations.

@@

\subsection{Monte Carlo Method Selection and Convergence}

@@

Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge
relative to the problem size. We see clearly in
Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint solver with
MCSA results in several orders of magnitude speedup over the direct
solver while the number of iterations required to converge is of a
similar scale. We expect this for several reasons. First, with an
equivalent number of histories specified for both solvers and a system
of size $N$, the direct solver will compute $N$ times the number of
histories random walks while the adjoint solver will only compute the
number of histories specified. Initially, this may not seem like a
fair comparison, however, we see from Fig.~\ref{fig:poisson_cpu_time}
that the number of iterations requried to converge is approximately
the same. This is necessary in the direct method to ensure a
contribution from each state as the random walk sequence will only
contribute to the starting state. For the adjoint method, because the
random sequence contributes to the state in which it currently
resides, fewer histories are neccessary to compute contributions from
all the nonzero parts of the source. Therefore, although the total
number of histories per cycle is in fact larger for the direct method,
the solution computed at each iteration is no better than that
computed by the adjoint method.

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.eps}
  \caption{\sl CPU Time (s) to converge vs. Problem Size (N for an NxN
    mesh). Both the adjoint and direct solvers are used with the five
    point and nine point stencils. A significant CPU time speedup is
    noted with the adjoint method due to the reduced number of
    histories.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.eps}
  \caption{\sl Iterations to converge vs. Problem Size (N for an NxN
    mesh. Both the adjoint and direct solvers are used with the five
    point and nine point stencils. Both methods give similar iteration
    behavior.}
  \label{fig:poisson_iterations}
\end{figure}

@@

The discrete diffusion equation in (\ref{eq:discrete-diffusion}) can
be written in matrix-operator form as follows,
\begin{equation}
  \ve{D}\bphi = \ve{q}\:,
  \label{eq:diffusion-matrix}
\end{equation}
where $\ve{D}$ is a symmetric, positive-definite (SPD) operator.
Unfortunately, for many problems $\rho(\ve{D}) > 1$.  Therefore, we
must precondition Eq.~(\ref{eq:diffusion-matrix}) in order to use the
MCSA method. We apply left-preconditioning and write
\begin{equation}
  \ve{M}^{-1}\ve{D}\bphi = \ve{M}^{-1}\ve{q}\:.
  \label{eq:preconditioned-diffusion}
\end{equation}
A good choice for the preconditioner, $\ve{M}$, is
\begin{equation}
  \ve{M}=\diag(\ve{D})\:.
  \label{eq:preconditioner}
\end{equation}
Setting $\ve{M}$ equal to the diagonal elements of $\ve{D}$
accomplishes several objectives:
\begin{itemize}
\item It reduces the spectral radius such that
  $\rho(\ve{M}^{-1}\ve{D}) < \rho(\ve{D})$ and
  $\rho(\ve{M}^{-1}\ve{D}) < 1$.  Thus, convergence is guaranteed and
  fewer iterations are required resulting in accelerated convergence.
\item The diagonal elements of the preconditioned iteration matrix are
  zero, $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no
  time wasted in transitions $i\rightarrow i$ during the random walk
  process.
\end{itemize}
The second item is quite important for Monte Carlo solutions of linear
systems, and this type of scaling should be performed regardless of
any additional preconditioner choices.

@@

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA
method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
method requires two solution steps:

\begin{enumerate}
\item A fixed-point iteration to estimate $\bphi^{l+1/2}$.  This is a
  matrix-vector multiply operation.
\item An adjoint Monte Carlo solve to estimate $\delta\bphi^{l+1/2}$.
\end{enumerate}

For clarity, we rewrite the iteration scheme in
Eqs.~(\ref{eq:general-MCSA}) using the diffusion operator notation
from Eq.~(\ref{eq:preconditioned-diffusion}),

\begin{align*}
  %%
  \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
  \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\
  %%
  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
  \quad & (\text{compute residual})\\
  %%
  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
  (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\
  %%
  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
  (\text{update}\ \bphi)
\end{align*}

As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo
method only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now
describe each of these steps in turn.

First, we use Eq.~(\ref{eq:preconditioned-discrete-diffusion}) to
evaluate $\bphi^{l+1/2}$ via a single fixed-point iteration as
follows:

\begin{equation}
  \begin{aligned}
    \phi_{ijk}^{l+1/2} =  
    \frac{\sigm_{i+1}}{\sigT}\frac{\Delta_{i+1}}{\Di}&\phi_{i+1\,jk}^l
    + \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}^l
    +
    \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}^l \\ 
    &+ \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}^l 
    + \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}^l
    + \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}^l
    + \frac{\qn_{ijk}}{\sigT}\:.
  \end{aligned}
\end{equation}

This result is used to compute the residual of the system.  The
residual then becomes the source for the Monte Carlo simulation.

@@

\begin{figure}[htpb!]
  \begin{center}
    \begin{tabular}{cc}
      \subfigure[Edit points]{\includegraphics[width=2in,clip]{points.eps}} &
      \subfigure[Temperature]{\includegraphics[width=4in,clip]{multi_mat.eps}}
    \end{tabular}
  \end{center}
  \caption{(a) Edit points and (b) time-evolution of the temperature
    at each point.  All points are centered in the $z$-axis.}
  \label{fig:edit-points}
\end{figure}
